Icicle: write once, run once

### 2016-05-12 2016fhpc/Main.tex
ok
introduction and first section (Source), start on Core section

\title{Icicle: fuse your queries}

@@

\begin{abstract}
When streaming a large amount of data, simply iterating over the data may take hours.
If multiple queries are to be performed, it is important that work is not duplicated.
Queries that can be performed together must be performed in the same iteration.
We introduce a simple streaming language for computing queries in a single-pass over the data.
By using an appropriate intermediate language we guarantee fusion between queries on the same input streams, before extracting efficient C code.
\end{abstract}

@@

\input{section/01-Introduction.tex}
\input{section/02-Core.tex}

>>>>

\title{Icicle: write once, run once}

@@

\begin{abstract}
When querying a large amount of data, simply iterating over the data may take hours.
If multiple queries are to be performed, it is important that work is not duplicated.
Queries that can be performed together must be performed in the same iteration.
We introduce a simple streaming language for computing queries in a single-pass over the data.
With type and other restrictions we guarantee fusion between queries on the same input table, before extracting efficient C code.
\end{abstract}

@@

\input{section/01-Introduction.tex}
\input{section/02-Source.tex}
\input{section/03-Core.tex}
\input{section/04-Benchmarks.tex}
\input{section/05-Conclusion.tex}
\input{section/XX-Elimination.tex}
\input{section/XX-Related.tex}

### 2016-05-12 2016fhpc/section/01-Introduction.tex
introduction and first section (Source), start on Core section

For example, suppose we have a table called @stocks@ with open and close prices for each company, for each day.
We may wish to find the number of days where the open price exceeds the close price, and vice versa:
\begin{code}
SELECT COUNT(*) AS open_gt_close
FROM   stocks
WHERE  open > close;

SELECT COUNT(*) AS open_lt_close
FROM   stocks
WHERE  open < close;
\end{code}

In order to fuse these so that both are computed in a single query, some ingenuity is required.
This is because the @WHERE@ clause filters over the entire query, but when putting both in one query we need two different filters.
\begin{code}
SELECT COUNT(IF(open > close, 1, 0)) AS open_gt_close
     , COUNT(IF(open < close, 1, 0)) AS open_lt_close
FROM stocks;
\end{code}

If we also wish to compute the mean of open price on those days, we can do this like so:
\begin{code}
SELECT SUM(open) / COUNT(*) AS mean
FROM   stocks
WHERE  open > close;
\end{code}

Now we can perform a similar trick to fuse all three together.
\begin{code}
SELECT COUNT(IF(open > close, 1, 0)) AS open_gt_close
     , COUNT(IF(open < close, 1, 0)) AS open_lt_close
     , SUM  (IF(open > close, open, 0))
     / COUNT(IF(open > close, 1, 0)) AS mean
FROM stocks;
\end{code}

@@

Another problem is that we have no guarantee that duplicate computations introduced by fusion will be eliminated.
Above we are computing the count where open exceeds close twice: once in @open_gt_close@ and again in @mean@.

The purpose of Icicle is two-fold: we wish to ensure that all queries over the same table can be fused together, and guarantee that such duplicate computations will be removed.
This is achieved by having a very restricted query language with no sorting, nested subqueries, and allowing only a single-pass over the data.

In Icicle, the above queries are expressed as:
\begin{code}
feature stocks
in filter open > close
in count.

feature stocks
in filter open < close
in count.

feature stocks
in filter open > close
in sum open / count.
\end{code}

These are then converted into the following query plans before fusion:
\begin{code}
query stocks {
  fold count  = 0
           then if open > close then count + 1 else count;
  output open_gt_close = count;
}
\end{code}

>>>>

For example, suppose we have a table called @stocks@ with open and close prices for each company, for each day.
We may wish to find the number of days where the open price exceeds the close price, and vice versa:
\begin{code}
SELECT COUNT(*)
FROM   stocks
WHERE  open > close;

SELECT COUNT(*)
FROM   stocks
WHERE  open < close;
\end{code}

In order to fuse these so that both are computed in a single query, some ingenuity is required.
This is because the @WHERE@ clause filters over the entire query, but when putting both in one query we need two different filters.
\begin{code}
SELECT COUNT(IF(open > close, 1, 0))
     , COUNT(IF(open < close, 1, 0))
FROM stocks;
\end{code}

If we also wish to compute the mean of open price on those days, we can do this like so:
\begin{code}
SELECT SUM(open) / COUNT(*)
FROM   stocks
WHERE  open > close;
\end{code}

Now we can perform a similar trick to fuse all three together.
\begin{code}
SELECT COUNT(IF(open > close, 1, 0))
     , COUNT(IF(open < close, 1, 0))
     , SUM  (IF(open > close, open, 0))
     / COUNT(IF(open > close, 1, 0))
FROM stocks;
\end{code}

@@

Another problem is that we have no guarantee that duplicate computations introduced by fusion will be eliminated.
Above we are computing the count where open exceeds close twice: once in the count, and again for the mean.

The purpose of Icicle is two-fold: we wish to ensure that all queries over the same table can be fused together, and guarantee that such duplicate computations will be removed.
This is achieved by having a very restricted query language with no sorting, and allowing only a single-pass over the data.

In Icicle, the above queries are expressed as:
\begin{code}
query stocks
in filter open > close
in count.

query stocks
in filter open < close
in count.

query stocks
in filter open > close
in sum open / count.
\end{code}

These queries are then fused together, common subexpression elimination is performed, and efficient C code is generated.

The contributions of this paper are:
\begin{itemize}
\item
we describe restrictions on Icicle which guarantee any two queries on the same table can be fused together\sref{s:IcicleSource};
\item
we show an intermediate fold-based language, which makes fusion a simple matter of appending two programs\sref{s:IcicleCore};
\item
and we present benchmarks of our generated code, which compares favourably against standard unix utilities and outperforms our existing system by orders of magnitude\sref{s:Benchmarks}.
\end{itemize}

### 2016-05-12 2016fhpc/section/03-Core.tex
introduction and first section (Source), start on Core section

%!TEX root = ../Main.tex
\section{Icicle Core}
\label{s:IcicleCore}

@@

\begin{code}
N     := a,b,...
FoldN := a,b,...

Exp   := N
       | FoldN  | new FoldN
       | Exp Exp
       | \N. Exp

Bind  := fold FoldN : T = Exp then Exp;
       | let  N     : T = Exp;

Stage := stage { Bind...  }
Query := loop N : T { Stage... }

\end{code}

\caption{Grammar}
\label{fig:grammar}
\end{figure}

\newcommand\TypecheckX[3]
{                   #1
        ~\vdash~    #2~:~#3
}

\newcommand\TypecheckS[3]
{                   #1
        ~\vdash~    #2~\dashv~#3
}

\begin{figure*}

$$
\boxed{\TypecheckS{\Gamma}{\mi{Query}}{\Gamma'}}
$$

$$
\ruleAx
{
  \TypecheckS{\Gamma}{@loop@~x~:~\tau~\{\}}{\emptyset}
}{QueryNil}
\ruleIN
{
  \TypecheckS{\Gamma,~x~:~\tau}{s}{\Gamma'}
  \quad
  \TypecheckS{\Gamma,~\Gamma'}{@loop@~x~:~\tau~\{s'\}}{\Gamma''}
}{
  \TypecheckS{\Gamma}{@loop@~x~:~\tau~\{s;~s'\}}{\Gamma' \cup \Gamma''}
}{QueryCons}
$$

$$
\boxed{\TypecheckS{\Gamma}{\mi{Stage}}{\Gamma'}}
$$

$$
\ruleAx
{
  \TypecheckS{\Gamma}{@stage@~\{\}}{\emptyset}
}{StageNil}
\ruleIN
{
  \TypecheckS{\Gamma,~x~:~\tau}{@stage@~\{s'\}}{\Gamma'}
  \quad
  \TypecheckX{\Gamma}{x}{\tau}
}
{
  \TypecheckS{\Gamma}{@stage@~\{@let@~x~:~\tau~=~e;~s'\}}{\Gamma',~x~:~\tau}
}{StageLet}
$$

$$
\begin{array}{l}

\ruleIN
{
  \TypecheckS{\Gamma,~x~:~\tau,~@new@~x~:~\tau}{@stage@~\{s'\}}{\Gamma'}

  \quad
  \TypecheckX{\Gamma}{z}{\tau}
  \quad
  \TypecheckX{\Gamma~\cup~\mi{folds}}{k}{\tau}
}
{
  \TypecheckS{\Gamma}{@stage@~\{@fold@~x~:~\tau~=~z~@then@~k;~s'\}}{\Gamma',~x~:~\tau,~@new@~x~:~\tau}
}{StageFold}

  \\
  \mbox{where} \\
  \quad
  \mi{folds}~=~\{v~:~\tau'~|~@new@~v~:~\tau'~\in~\Gamma'\}
  \end{array}
$$

\TODO{This is wrong: need two environments}

\caption{Typing rules}
\label{fig:typing}
\end{figure*}

The full grammar is in figure~\ref{fig:grammar}, however primitives and types have been omitted.
It is assumed that there are some primitives such as @1@, @+@ and types such as @Int@, but the elements do not actually matter.

The typing rules are in figure~\ref{fig:typing}.
The rules shown only cover the staging and folds.
As with the grammar, we assume there is a typing rule $\TypecheckX{\Gamma}{\mi{Exp}}{\tau}$ which checks an expression under the given environment.

The first judgment, $\TypecheckS{\Gamma}{\mi{Query}}{\Gamma'}$

>>>>

%!TEX root = ../Main.tex
\section{Intermediate language}
\label{s:IcicleCore}

\begin{code}
Exp   := x | Prim Exp...

Plan  := query  x : T
       before { x : T = Exp... }
       folds  { x : T = Exp then Exp... }
       after  { x : T = Exp... }
       return { x : T = x... }

T     := bool | int | map T T
\end{code}

\caption{Grammar}
\label{fig:core:grammar}
\end{figure}

The intermediate language can be thought of as similar to a query plan in other databases.
Queries in the source language are converted to query plans, and then query plans on the same table are fused together.
Common subexpression elimination, partial evaluation and other optimisations are performed at the query plan level.

The grammar for the intermediate language is given in figure~\ref{fig:core:grammar}.
Expressions can be variables or primitive applications: function applications are not allowed in expressions here, as their definitions are inlined before converting to query plans.
The @Plan@ itself is split into stages of computation, which correspond to the modal stages in the Source query.
The first stage, @before@, is used for pure computations which do not depend on any stream values at all.
The next stage, @folds@, defines the @Element@ computations and how they are converted into @Aggregate@ computations.
Next are @after@ which correspond to @Aggregate@ computations, occurring after the entire stream has been seen.
Finally, the @return@ specifies the output values of the query; a single query will have only one output value, but the result of fusion can have many outputs.

@@

### 2016-05-12 2016fhpc/section/04-Benchmarks.tex
introduction and first section (Source), start on Core section

\begin{figure*}
$$\begin{array}{c}

\begin{tabular}{lrrrrrrrr}
                & \multicolumn{2}{c}{Unfused}         & \multicolumn{2}{c}{Stream}
                & \multicolumn{2}{c}{Megiddo} &\multicolumn{2}{c}{\textbf{Ours}} \\
                & Time & Loops   & Time & Loops      & Time & Loops & Time & Loops   \\
\hline
Normalize2      & 1.88s & 5      & 1.64s & 4          & 1.82s & 3  & \textbf{1.59s} & \textbf{2}\\
Closest points  & 3.83s & 6      & 3.33s & 5          & 2.92s & 3  & \textbf{2.92s} & \textbf{3}\\
QuadTree        & 5.22s & 8      & 5.22s & 8          & 4.72s & 2  & \textbf{4.72s} & \textbf{2}\\
\end{tabular}

\end{array}$$
\caption{Benchmark results}
\label{f:BenchResults}
\end{figure*}

This section discusses three representative benchmarks, and gives the full ILP program of the first. These benchmarks highlight the main differences between our fusion mechanism and related work. The runtimes of each benchmark are summarized in Figure~\ref{f:BenchResults}. We report times for: the unfused case where each operator is assigned to its own cluster; the clustering implied by stream fusion~\cite{coutts2007streamfusion}; the clustering chosen by Megiddo~\cite{megiddo1998optimal}, and the clustering chosen by our system.

For each benchmark we report the runtimes of hand-fused C code based on the clustering determined by each algorithm. Although we also have an implementation of our Data Flow Fusion system in terms of a GHC plugin~\cite{lippmeier2013flow}, we report on hand-fused C code to provide a fair comparison to related work. As mentioned in~\cite{lippmeier2013flow}, the current Haskell stream fusion mechanism introduces overhead in terms of a large number of duplicate loop counters, which increases register pressure unnecessarily. Hand fusing all code and compiling it with the same compiler (GCC) isolates the true cost of the various clusterings from low level differences in code generation.

We have used both GLPK and CPLEX as external ILP solvers. For small programs such as @normalizeInc@, both solvers produce solutions in under 100ms. For a larger randomly generated example with twenty-five combinators, GLPK took over twenty minutes to produce a solution while the commercial CPLEX solver was able to produce a solution in under one second --- which is still quite usable. We will investigate the reason for this wide range in performance in future work.

% We use GLPK as an external solver, which is not particularly fast, but has readily-available Haskell bindings. For these small programs with few combinators, it produces results in under one hundred milliseconds.
% For larger programs with twenty or more combinators to be clustered, GLPK is inadequate.
% One randomly generated example with twenty-five combinators took over twenty minutes in GLPK, while the commercial solver CPLEX was able to solve the same example in a second, and another open source solver CBC took one hundred seconds.
% For the same example, the unoptimised program with all constraints took over twenty minutes in CBC, while CPLEX still only took one and a half seconds.

The benchmark programs are at \url{https://github.com/amosr/papers/tree/master/2014betterfusionforfilters/benches}.

% The three benchmarks are the @normalize2@ example, finding the closest pair of points, and quadtree. The benchmark programs are all hand-written, hand-fused C code based on the clustering. Each program was run with the same input five times, and the minimum runtime was used. Runtimes and the number of loops for each clustering are shown in Figure~\ref{f:BenchResults}. In all cases, our clustering performs better than or as good as Megiddo's, and better than stream fusion and unfused. Interestingly, stream fusion's clustering for @normalize2@ performs better than Megiddo's, despite having more loops, as stream fusion is able to remove the intermediate array.

% -----------------------------------------------------------------------------
\pagebreak
\subsection{Normalize2}
To demonstrate the ILP formulation we will use the @normalize2@ example from \S\ref{s:Introduction}, repeated here:
\begin{code}
  normalize2 :: Array Int -> Array Int
  normalize2 xs
   = let sum1 = fold   (+)  0   xs
         gts  = filter (>   0)  xs
         sum2 = fold   (+)  0   gts
         ys1  = map    (/ sum1) xs
         ys2  = map    (/ sum2) xs
     in (ys1, ys2)
\end{code}

We use the final ILP formulation from \S\ref{s:OptimisedConstraints}. First, we calculate $possible$ -- that is, the nodes which have no fusion-preventing path between them.
\[ \{ \{sum1, gts, sum2\}
 , \{sum1, ys2\}
 , \{gts, sum2, ys1\}
 , \{ys1, ys2\} \} \]

The complete ILP program is shown below. Note that in the objective function the weights for $x_{sum1, sum2}$ and $x_{sum2, ys1}$ are both only 1, because they do not share any input arrays.

\begin{tabbing}
MMMMM   \= MMMMMMM \= M \= MMMMMMM \= M \= MMMMMMM \= \kill
Minimise   \> $25 \cdot x_{sum1, gts} + 1  \cdot x_{sum1,sum2} + 25 \cdot x_{sum1, ys2} +$ \\
           \> $25 \cdot x_{gts, sum2} + 25 \cdot x_{gts, ys1} + 1 \cdot x_{sum2, ys1} +$ \\
           \> $25 \cdot x_{ys1, ys2}  + 5  \cdot c_{gts} + 5 \cdot c_{ys1} + 5 \cdot c_{ys2} $
\\[0.5ex]
Subject to 
    \> $-5 \cdot x_{sum1, gts}$  \> $\le$ \> $\pi_{gts} - \pi_{sum1}$  \> $\le$ \> $5 \cdot x_{sum1, gts}$  \\
    \> $-5 \cdot x_{sum1, sum2}$ \> $\le$ \> $\pi_{sum2} - \pi_{sum1}$ \> $\le$ \> $5 \cdot x_{sum1, sum2}$ \\
    \> $-5 \cdot x_{sum1, ys2 }$ \> $\le$ \> $\pi_{ys2 } - \pi_{sum1}$ \> $\le$ \> $5 \cdot x_{sum1, ys2 }$ \\
    \> $-5 \cdot x_{gts,  ys1 }$ \> $\le$ \> $\pi_{ys1 } - \pi_{gts }$ \> $\le$ \> $5 \cdot x_{gts, ys1  }$ \\
    \> $-5 \cdot x_{sum2, ys1 }$ \> $\le$ \> $\pi_{ys1 } - \pi_{sum2}$ \> $\le$ \> $5 \cdot x_{sum2, ys1 }$ \\
    \> $-5 \cdot x_{ys1, ys2  }$ \> $\le$ \> $\pi_{ys2 } - \pi_{ys1 }$ \> $\le$ \> $5 \cdot x_{ys1, ys2  }$ 
\\[0.5ex]
    \> $   x_{gts, sum2 }$ \> $\le$ \> $\pi_{sum2} - \pi_{gts }$ \> $\le$ \> $5 \cdot x_{gts, sum2 }$ 
\\[0.5ex]
    \>                     \>       \> $\pi_{sum1} < \pi_{ys1}$ \\
    \>                     \>       \> $\pi_{sum2} < \pi_{ys2}$
\\[0.5ex]
    \> $ x_{gts,sum2} $    \> $\le$ \> $c_{gts}$
\\[0.5ex]
    \> $x_{gts, sum2}$     \> $\le$ \> $x_{sum1, sum2}$ \\
    \> $x_{sum1,sum1}$     \> $\le$ \> $x_{sum1, sum2}$ \\
    \> $x_{sum1, gts}$     \> $\le$ \> $x_{sum1, sum2}$
\end{tabbing}
\noindent

One minimal solution to this is:
\begin{tabbing}
MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
$x_{sum1, gts},~ x_{sum1, sum1},~ x_{sum1, sum2},~ x_{gts, sum2},~ x_{ys1,  ys2}$
    \> $=$ \> $0$ \\
$x_{sum1, ys2},~ x_{gts, ys1},~   x_{sum2, ys1}$
    \> $=$ \> $1$ 
\\[1ex]
$\pi_{sum1},~ \pi_{gts },~ \pi_{sum2}$
    \> $=$ \> $0$ \\
$\pi_{ys1 },~ \pi_{ys2 }$
    \> $=$ \> $1$ 
\\[1ex]
$c_{gts},~ c_{ys1},~ c_{ys2}$           
    \> $=$ \> $0$
\end{tabbing}

This minimal solution is not unique, though in this case the only other minimal solutions use different $\pi$ values, and denote the same clustering. Looking at just the non-zero variables in the objective function, the value is $25 \cdot x_{sum1,ys2} + 25 \cdot x_{gts,ys1} + 1 \cdot x_{sum2, ys1} = 51$. For illustrative purposes, note that objective function could be reduced by setting $x_{sum1,ys2} = 0$ (fusing $sum1$ and $ys1$), but this conflicts with other constraints. Since $x_{sum1, sum2} = 0$, we require $\pi_{sum1} = \pi_{sum2}$, but also \mbox{$\pi_{sum2} < \pi_{ys2}$}. These constraints cannot be satisfied, so a clustering that fused $sum1$ and $ys2$ would not also permit $sum1$ and $sum2$ to be fused.

We will now compare the clustering produced by our system, with the one implied by stream fusion. As described in \cite{lippmeier2013flow}, stream fusion cannot fuse a produced array into multiple consumers, or fuse operators that are not in a producer-consumer relationship. The corresponding values of the $x_{ij}$ variables are:
\begin{tabbing}
MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
$x_{gts, sum2}$
    \> $=$ \> $0$ \\
$x_{sum1, gts}, x_{sum1, sum2}, x_{ys1,  ys2}, x_{sum1, ys2}, x_{gts, ys1 }, x_{sum2, ys1}$
    \> $=$ \> $1$
\end{tabbing}

We can force this clustering to be applied in our integer linear program by adding the above equations as new constraints. Solving the resulting program then yields:
\begin{tabbing}
MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
$\pi_{sum1}, \pi_{gts }, \pi_{sum2}$
    \> $=$ \> $0$ \\
$\pi_{ys1 }, \pi_{ys2 }$
    \> $=$ \> $1$ \\
$c_{gts}, c_{ys1}, c_{ys2}$           
    \> $=$ \> $0$
\end{tabbing}

Note that although nodes $sum1$ and $sum2$ have equal $\pi$ values, they are not fused. Conversely, if two nodes have different $\pi$ values then they are never fused.

% The corresponding value of the objective function is: $25 \cdot x_{sum1, gts} + 1 \cdot x_{sum1,sum2} + 25 \cdot x_{sum1, ys2} + 25 \cdot x_{gts, ys1} + 1 \cdot x_{sum2, ys1} + 25 \cdot x_{ys1, ys2} = 102$.

% -----------------------------------------------------------------------------
\subsection{Closest Points}
The closest points benchmark is a divide-and-conquer algorithm that finds the closest pair of 2-dimensional points in an array. We first find the midpoint along the Y-axis, and filter the remaining points to those above and below the midpoint. We then recursively find the closest pair of points in the two halves, and merge the results. As the filtered points are passed directly to the recursive call, there is no further opportunity to fuse them, and our clustering is the same as returned by Megiddo's algorithm. However, our clustering generates both filtered arrays in a single loop, unlike stream fusion that requires a separate loop for each.

% -----------------------------------------------------------------------------
\subsection{QuadTree}
The QuadTree benchmark recursively builds a 2-dimensional space partitioning tree from an array of points. At each step the array of points is filtered into four 2-dimensional boxes. As with the closest points algorith, there are no further opportunities for fusing the filtered results, and our clustering is the same as Megiddo's. However, our clustering produces all four filtered results in a single loop, whereas stream fusion requires four loops.

% -----------------------------------------------------------------------------
\subsection{QuickHull}
The core of the QuickHull algorithm is shown below: given a line and an array of points, we filter the points to those above the line, and also find the point farthest from that line.

\begin{code}
hull :: (Point,Point) -> Array Point -> Array Point
hull line@(l,r) pts
 = let pts' = filter (above   line) pts
       ma   = fold   (maxFrom line) pts'
   in (hull (l, ma) pts') ++ (hull (ma, r) pts')
\end{code}

Stream fusion cannot fuse the @pts'@ and @ma@ bindings because @pts'@ is referred to multiple times and thus cannot be inlined. Megiddo's algorithm also cannot fuse the two bindings because their iteration sizes are different. If the @ma@ binding was rewritten to operate over the @pts@ array instead of @pts'@, Megiddo's formulation would be able to fuse the two, and the overall program would give the same result. However, this performance behavior is counter intuitive because @pts'@ is likely to be smaller than @pts@, so in an unfused program the original version would be faster. Our system fuses both versions.

>>>>

Run some benchmarks.

### 2016-05-12 2016fhpc/section/XX-Elimination.tex
introduction and first section (Source), start on Core section

@@

We can look at the topological ordering of stages as layers of the same depth.
In the @count/sum/ab@ example, @count@ and @sum@ are together on the top layer, while @ab@ is on the second layer.

Each stage need only be compared with those on the same layer, and those on the layer directly above.
In fact, of those on the layer above, it only needs to be compared with those it refers to.
\TODO{Justify this further.}
Because the stages have already been split apart, we also need only check against those with equal or greater number of bindings.

\subsection{Duplicate removal}
Because the stages have already been split apart, we can only replace a stage

>>>>

\subsection{Stage-local duplicate removal}
Here is an example of a stage with two counters.
\begin{code}
stage {
  fold a = 0 then c + 1
  fold b = 0 then c + 1
  fold c = 0 then a + b
}
Execution:
[ a = 0, b = 0, c = 0 ]
[ a = 1, b = 1, c = 0 ]
[ a = 1, b = 1, c = 2 ]
[ a = 3, b = 3, c = 2 ]
[ a = 3, b = 3, c = 6 ]
[ a = 7, b = 7, c = 6 ]
[ a = 7, b = 7, c = 14 ]
\end{code}

The bindings for @a@ and @b@ here are identical, and so one can be replaced with the other, by substituting @b := a@ and removing the binding for @b@.
\begin{code}
stage {
  fold a = 0 then c + 1
  fold c = 0 then a + a
}
\end{code}

Another example is a single count, spread across two recursive bindings.
Here, @a@ and @b@ depend on each other, but both have the same body modulo alpha.
\begin{code}
stage {
  fold a = 0 then b + 1
  fold b = 0 then a + 1
}
Execution:
[ a = 0, b = 0 ]
[ a = 1, b = 1 ]
[ a = 2, b = 2 ]
\end{code}

One can place them into equivalence groups based on their bodies: @a@ and @b@ go in the same equivalence group, and so any reference to @a@ or @b@ refers to the equivalence group containing both.
Equivalence groups are then squashed together, with a single binding for each equivalence group.
\begin{code}
stage {
  fold a = 0 then a + 1
}
\end{code}

@@

We can look at the topological ordering of stages as layers of the same depth.
In the @count/sum/ab@ example, @count@ and @sum@ are together on the top layer, while the stage containing @a@ and @b@ is on the second layer.

When checking for duplicates, each stage need only be compared with those on the same layer, and those on the layer directly above.
In fact, of those on the layer above, it only needs to be compared with those it refers to.
It is not necessary to check more two layers above, for example, because we know two things:
this stage refers to a binding in the layer above, while any stage two layers above cannot possibly refer to a lower stage.
However, in order to be duplicates, they must refer to the same things, or duplicates of the same things.

\subsection{Inter-stage duplicate removal}
To check if one stage is a duplicate of another, we go through each binding of the first stage, checking if there is an alpha-equivalent binding in the other stage.
We also create a substitution between the two stages.
This substitution is later performed on the rest of the program, so that after the duplicate is removed, references to the duplicate are updated to refer to the first stage.

During the process of checking, there may be multiple possible substitutions that appear to work.
The following program defines three mutually recursive counters; two of the counters increment by one, and the last increments by two.
Because of the mutual recursion, this ends up making each counter increment by a repeating pattern: one, one, two.
\begin{code}
stage {
  fold a = 0 then b + 1
  fold b = 0 then c + 1
  fold c = 0 then a + 2
}
Execution:
[ a = 0, b = 0, c = 0 ]
[ a = 1, b = 1, c = 2 ]
[ a = 2, b = 3, c = 3 ]
[ a = 4, b = 4, c = 4 ]
[ a = 5, b = 5, c = 6 ]
[ a = 6, b = 7, c = 7 ]
\end{code}

Now, suppose we wish to check if the following stage is a duplicate of the previous counters.
Note that these bindings are actually in a different order to the previous bindings, whereas the equivalent order would be @x@, @y@ then @z@.
\begin{code}
stage {
  fold y = 0 then z + 1
  fold x = 0 then y + 1
  fold z = 0 then x + 2
}
\end{code}

We start by inpsecting @y@, the first binding, and check against all bindings in the other set.
We wish to find a substitution from the names @x@, @y@ and @z@ to the names @a@, @b@ and @c@.
The substitution cannot mention any names other than those bound by the two stages.

There appear to be two possibilities for @y@: it is alpha-equivalent to both @a@ and @b@, producing two different substitutions:
\begin{code}
(Y1)
y := a
z := b
(Y2)
y := b
z := c
\end{code}

We continue checking the remaining bindings, by producing substitutions for @x@.
Like @y@, @x@ fits both @a@ and @b@, with two possible substitutions:
\begin{code}
(X1)
x := a
y := b
(X2)
x := b
z := c
\end{code}

We can cross-product these substitutions together, composing (Y1) with (X1) and so on, producing four possible substitutions.
The first two contain contradictions and are discarded immediately.
The next two seem plausible, so far.
\begin{code}
(Y1X1)
y := a
y := b
(Y1X2)
z := b
z := c
(Y2X1)
x := a
y := b
z := c
(Y2X2)
x := b
y := b
z := c
\end{code}

Now we check the final binding, @z@, which produces only one possible substitution:
\begin{code}
(Z1)
z := c
x := a
\end{code}

Composing (Z1) with (Y2X1) and (Y2X2), we find that (Y2X1Z1) is the only remaining possibility, since (Z1) and (Y2X2) are contradictory.
\begin{code}
(Y2X1Z1)
x := a
y := b
z := c
(Y2X2Z1)
x := a
x := b
\end{code}

We can now remove the stage containing @x@, @y@ and @z@, while performing this substitution over the remainder of the program, so that any references to @x@ become references to @a@, and so on.

If, for a different pair of stages, there were multiple possible substitutions after checking all bindings against the others, we could have chosen any of the substitutions.
They are all equivalent.
